Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2015-09-05 pkg/glmertree/paper/treatment_interactions.Rnw
major revision of the glmertree package -> arguments and return values renamed to better match other functions -> streamlined code (e.g., special-casing of root-only tree should not    be necessary anymore) -> added S3 class with some basic methods -> collected example data in a single object with (somewhat) better name -> adapted paper accordingly
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@774 255f4249-2263-4417-b1ea-8498779e27a7

<<example-trees>>=
exampleglm <- lmtree(outcome ~ treatment | age, data=example)
exampleglmtree <- lmtree(outcome ~ treatment | age + duration + anxiety, data=example)
exampleglmmtree <- lmertree(outcome ~ treatment | age + duration + anxiety, outcome ~ (1|cluster), data=example)
@

@@

\begin{figure}[!h]
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree$Tree)
@
\caption{GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

@@

\begin{figure}[!thbp]
<<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
load("treesizes_long_nointeract.dat")
treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered=T)
treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered=T)
aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, 
	type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
	abline=c(1,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1, 
	lines=TRUE, points=TRUE)))
@
\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=1$ represents the true tree size.}
\label{fig:xyplot_treesize_nointeract}
\end{figure}

@@

## GLMM tree
lmertree_app <- lmertree(formula, HRSDt1 ~ (1 | studyid) + HRSDt0, data = metadata,
 		initialRandomEffects = fitted(lm(HRSDt1 ~ HRSDt0, data = metadata)))
@

@@

\begin{figure}[!th]
\setkeys{Gin}{width=0.6\textwidth}
<<depression-glmmtree, fig=TRUE, height=4, width=4.7>>=
plot(lmertree_app$Tree)
@
\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
\label{fig:lmertree_C&W}
\end{figure}

>>>>

<<example-trees>>=
exampleglm <- lmtree(outcome ~ treatment | age, data = example)
exampleglmtree <- lmtree(outcome ~ treatment | age + duration + anxiety, data = example)
exampleglmmtree <- lmertree(outcome ~ treatment | age + duration + anxiety, outcome ~ (1|cluster), data = example)
@

@@

\begin{figure}[!h]
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree)
@
\caption{GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

@@

\begin{figure}[!thbp]
<<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
load("treesizes_long_nointeract.dat")
treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered = TRUE)
treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered = TRUE)
aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, 
	type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
	abline=c(1,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1, 
	lines=TRUE, points=TRUE)))
@
\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=1$ represents the true tree size.}
\label{fig:xyplot_treesize_nointeract}
\end{figure}

@@

## GLMM tree
lmertree_app <- lmertree(formula, HRSDt1 ~ (1 | studyid) + HRSDt0, data = metadata,
  ranefstart = fitted(lm(HRSDt1 ~ HRSDt0, data = metadata)))
@

@@

\begin{figure}[!th]
\setkeys{Gin}{width=0.6\textwidth}
<<depression-glmmtree, fig=TRUE, height=4, width=4.7>>=
plot(lmertree_app)
@
\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
\label{fig:lmertree_C&W}
\end{figure}
